Reproducible Research Assignment 1 - Quantifiable Self
========================================================

## Loading and preprocessing the data

```{r, echo=TRUE}
data <- read.csv("activity.csv", header=T)       ## Reading in the data
str(data)
data$date <- as.Date(data$date)                  ## Convert variable into class "Date"
```

## What is mean total number of steps taken per day?

Aggregating the data to total steps per day, while ignoring the missing values: 
```{r, echo=TRUE}
totalsteps <- aggregate(data$steps, by=list(date=data$date), sum)
colnames(totalsteps) <- c("date", "totalsteps")
head(totalsteps)
```

Making a histogram of the total number of steps taken each day: 

```{r, echo=T, fig.width=10}
library(ggplot2)
qplot(totalsteps, data=totalsteps)+geom_histogram(fill="grey", color="black")+labs(title="Total Number of Steps Taken per Day", x="Steps per Day", y="Frequency")
```

Calculate the **mean** total number of steps taken per day: 
```{r, echo=TRUE}
mean(totalsteps$totalsteps, na.rm=T)
```
The mean of total number of steps per day is `r mean(totalsteps$totalsteps, na.rm=T)`. 

And the **median** total number of steps taken per day: 
```{r, echo=TRUE}
median(totalsteps$totalsteps, na.rm=T)
```
The median is `r median(totalsteps$totalsteps, na.rm=T)`. 

## What is the average daily activity pattern?

Make a time series plot:
```{r, echo=TRUE, fig.width=10}
library(plyr)
totalstepsinterval <- ddply(data, .(interval), summarise, mean_steps=mean(steps, na.rm=TRUE))
head(totalstepsinterval)
plot(totalstepsinterval, type="l", main="Average number of steps taken per 5-minute interval", xlab="5-minute intervals", ylab="Average number of steps (across all days)")
```

Display the 5-minute interval that contains the maximum number of steps: 
```{r, echo=TRUE}
max <- totalstepsinterval[which.max(totalstepsinterval$mean_steps),]
max 
```
The maximum number of steps are in the `r max[,1]`th interval. 


## Imputing missing values

Report the total number of missing values in the dataset: 
```{r, echo=TRUE}
sum(is.na(data$steps))
```
There are `r sum(is.na(data$steps))` rows with missing values. 

The imputation method chosen here is to replace the missing value by the average steps taken that day. 
```{r, echo=TRUE}
data_imputed <- data
data_imputed$steps[is.na(data_imputed$steps)]=mean(data_imputed$steps, by=list(interval), na.rm=TRUE)
data_imputed$steps[is.na(data_imputed$steps)] <- tapply(data$steps, data$interval, mean, na.rm=T)
#data_imputation <- merge(data_imputed, totalstepsinterval, by.x="interval", by.y="interval")
#head(data_imputation)
#data_imputation[is.na(data_imputation$steps),]$steps <- data_imputation[is.na(data_imputation$steps), 4]
#meanday <- ddply(data_imputed, .(date), summarise, mean_stepsday=mean(steps, na.rm=TRUE))

#data_imputed$steps[is.na(data$steps)] <- totalstepsinterval
#if(is.na(data$steps)==TRUE) {data$steps <- meanday}
totalsteps_imputed <- aggregate(data_imputed$steps, by=list(date=data_imputed$date), sum)
colnames(totalsteps_imputed) <- c("date", "totalsteps")
#head(totalsteps_imputed)
qplot(totalsteps, data=totalsteps_imputed)+geom_histogram(fill="grey", color="black")+labs(title="Total Number of Steps Taken per Day", x="Steps per Day", y="Frequency")
qplot(totalsteps_imputed, data=totalsteps_imputed)+geom_histogram(fill="grey", color="black")+labs(title="Total Number of Steps Taken per Day", x="Steps per Day", y="Frequency")


#df = data.frame(x = 1:20, y = c(1:10,rep(NA,10)))
#df$y[is.na(df$y)] = mean(df$y, na.rm=TRUE)
```

Calculate the **mean** total number of steps taken per day: 
```{r, echo=TRUE}
mean(totalsteps_imputed$totalsteps, na.rm=T)
```
The mean of total number of steps per day is `r mean(totalsteps_imputed$totalsteps, na.rm=T)`. 

And the **median** total number of steps taken per day: 
```{r, echo=TRUE}
median(totalsteps_imputed$totalsteps, na.rm=T)
```
The median is `r median(totalsteps_imputed$totalsteps, na.rm=T)`