---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Reproducible Research Assignment 1 - Quantifiable Self
========================================================


## Loading and preprocessing the data
```{r, echo=TRUE}
data <- read.csv("activity.csv", header=T)       ## Reading in the data
str(data)
data$date <- as.Date(data$date)                  ## Convert variable into class "Date"
```



## What is mean total number of steps taken per day?
Aggregating the data to total steps per day, while ignoring the missing values: 
```{r, echo=TRUE}
totalsteps <- aggregate(data$steps, by=list(date=data$date), sum)
colnames(totalsteps) <- c("date", "totalsteps")
head(totalsteps)
```

Making a **histogram** of the total number of steps taken each day: 

```{r, echo=T, fig.width=10}
library(ggplot2)
qplot(totalsteps, data=totalsteps)+geom_histogram(fill="grey", color="black")+labs(title="Total Number of Steps Taken per Day", x="Steps per Day", y="Frequency")
```

Calculate the **mean** total number of steps taken per day: 
```{r, echo=TRUE}
mean(totalsteps$totalsteps, na.rm=T)
```
The mean of total number of steps per day is `r mean(totalsteps$totalsteps, na.rm=T)`. 

And the **median** total number of steps taken per day: 
```{r, echo=TRUE}
median(totalsteps$totalsteps, na.rm=T)
```
The median is `r median(totalsteps$totalsteps, na.rm=T)`. 



## What is the average daily activity pattern?

Making a **time series plot**:
```{r, echo=TRUE, fig.width=10}
library(plyr)
totalstepsinterval <- ddply(data, .(interval), summarise, mean_steps=mean(steps, na.rm=TRUE))
head(totalstepsinterval)
plot(totalstepsinterval, type="l", main="Average number of steps taken per 5-minute interval", xlab="5-minute intervals", ylab="Average number of steps (across all days)")
```

Display the 5-minute interval that contains the maximum number of steps: 
```{r, echo=TRUE}
max <- totalstepsinterval[which.max(totalstepsinterval$mean_steps),]
max 
```
The maximum number of steps are in the `r max[,1]`th interval. 


## Imputing missing values

Report the total number of missing values in the dataset: 
```{r, echo=TRUE}
sum(is.na(data$steps))
```
There are `r sum(is.na(data$steps))` rows with missing values. 

The imputation method chosen here is to replace the missing value by the average of that 5-minute interval.
```{r, echo=TRUE}
totalstepsinterval <- ddply(data, .(interval), summarise, mean_steps=mean(steps, na.rm=TRUE))
```

Creating a new dataset with imputed values:
First, merge the data with missing values with the means per 5-minute interval. 
```{r, echo=TRUE}
data_imputed <- data
data_imputed <- merge(data_imputed, totalstepsinterval, by="interval")
```
Next, replace the NA values in a 5-minute interval by the corresponding mean value.
```{r, echo=TRUE}
for(i in 1:nrow(data_imputed)) {
    
    if(is.na(data_imputed$steps[i])){
        data_imputed$steps[i] <- data_imputed$mean_steps[i]
    }
}    
```
Verifying that there are no remaining missing values. No missing values anymore, so imputation worked. 
```{r, echo=TRUE}
sum(is.na(data_imputed$steps))
```



Making a **histogram** of the total number of steps taken each day: 
```{r, echo=TRUE}
newtotalsteps <- aggregate(data_imputed$steps, by=list(date=data_imputed$date), sum)
colnames(newtotalsteps) <- c("date", "totalsteps")
head(newtotalsteps)
qplot(totalsteps, data=newtotalsteps)+geom_histogram(fill="grey", color="black")+labs(title="Total Number of Steps Taken per Day", x="Steps per Day", y="Frequency")
```

Calculate the **mean** total number of steps taken per day: 
```{r, echo=TRUE}
mean(newtotalsteps$totalsteps, na.rm=T)
```
The mean of total number of steps per day is `r mean(newtotalsteps$totalsteps, na.rm=T)`. 

And the **median** total number of steps taken per day: 
```{r, echo=TRUE}
median(newtotalsteps$totalsteps, na.rm=T)
```
The median is `r median(newtotalsteps$totalsteps, na.rm=T)`.

Initially the mean was `r mean(totalsteps$totalsteps, na.rm=T)`and the median was `r median(totalsteps$totalsteps, na.rm=T)`. This implies that the new mean, after imputation, is still the same and the median has only changed by 1 step. 
After imputation there is less variation in the data distribution because the missing values were replaced by overall averages. The new median is now equal to the mean, the data appears to be more centered. 

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
Sys.setlocale("LC_TIME", "English")                     ## Ensuring that the display language is in English

data_imputed$weekday <- weekdays(data_imputed$date)                         ## Introducing weekday-variable
data_imputed$weekend <- as.factor(data_imputed$weekday %in% c("Saturday", "Sunday"))     ## Factor variable
levels(data_imputed$weekend) <- c("weekday", "weekend")                     ## weekday/weekend labels added
```

Make a **panel plot** for weekdays and weekends: 

Aggregating the data to averages per weekend or weekdays: 
```{r, echo=TRUE}
totalstepsweekend <- ddply(data_imputed, .(interval, weekend), summarise, avg_steps=mean(steps, na.rm=TRUE))
```
And then plotting the time series: 
```{r, echo=TRUE}
library(lattice)

xyplot(avg_steps ~ interval | weekend, data=totalstepsweekend, layout=c(1,2), type="l", xlab="5-minute interval", ylab="Average number of steps taken", main="Average Steps taken per 5-min interval - weekdays vs weekend")
```